Reasoning About a Machine with Local Capabilities

### 2018-04-15 presentations/esop18-04-2018/main.tex
more changes notes added to presentation

\begin{frame}
  \frametitle{Worlds, Safe Values, and Step-Indexing}
\begin{columns}
  \begin{column}{0.5\textwidth}
    \begin{itemize}
      \note[item]<1->{Compared to standard assembly language, capabilities executing code has access to represent bound.}
    \item<1-> Capabilities represent bounds on executing code
      \note[item]<1->{That is, the capabilities the executing code has access to.}
    \item<2-> World, $W$      
      \begin{itemize}
      \item Collection of invariants
      \end{itemize}
    \item<4-> Predicate for safe values w.r.t world, $\stdvr(W)$
      \begin{itemize}[<8->]
      \item Recursively defined
      \end{itemize}
    \end{itemize}
\end{column}
\begin{column}{0.5\textwidth}
  \begin{tikzpicture}[scale=.5, every node={scale=.5}]
    \scope
      \clip (-.1,-.1) rectangle (6.1,15.1);
      \fill[fill=white] (0,0) rectangle (6,15);
      \draw (0,0) -- (0,15);
      \draw (6,0) -- (6,15);
    \endscope
    % \draw[->] (-2,0) -- node[midway,sloped,above] {stack grows upward} (-2,15);
    \draw (-1.5,0) node {};
    \draw (-1.5,15) node {};
    \foreach \x in {0.5,1,...,14}
    {
      \draw[fill=white] (0,\x) rectangle (6,\x+.5) node[pos=.5,color=teal] {};
    };

    % One capability
    \draw<1-6> [decorate,decoration={brace,amplitude=3pt,mirror,raise=-2pt},yshift=0pt]
    (6.25,2) -- (6.25,5) node[draw=black] (sp1) [black,midway,xshift=6pt] {}; % ,shape=coordinate,xshift=1pt
    \draw<1-4>  (6,7.25) edge[black,->,in=45] (sp1);

@@

    \fill<3-6>[black, fill opacity=0, text opacity=0.7] (0,4) rectangle (6,4.5) node[pos=.5] {\footnotesize 42};

    \draw<5> (6,7.25) edge[black,->,in=45] node[right] {\footnotesize \textsc{r}} (sp1)   ;
    \draw<6> (6,7.25) edge[black,->,in=45] node[right] {\footnotesize \textsc{rw}} (sp1)   ;
    \draw<6>[very thick,color=red] (6.5,6.5) -- (8.5,4.5)
    (6.5,4.5) -- (8.5,6.5);

    \draw<7-> [decorate,decoration={brace,amplitude=3pt,mirror,raise=-2pt},yshift=0pt]
    (6.25,2) -- (6.25,5) node[draw=black] (sp1) [black,midway,xshift=6pt] {}; % ,shape=coordinate,xshift=1pt
    \path<7->  (6,4.25) edge[black,out=30,in=135] node[above right]  {\footnotesize \textsc{r}} (8,5) 
              (8,5) edge[black,->,out=-45,in=45](sp1);

@@

  \begin{overprint}
\onslide<1-4>
  \begin{itemize}
  \item<1-> Memory evolves over time
  \item<3-> Add invariants in future worlds
  \item<4-> Invariants as state machines
  \end{itemize}
  \onslide<5-6>
  \begin{itemize}
  \item<5-> Each state contains a predicate of accepted memory segments
  \item<6-> World indexed
  \end{itemize}
  \[
    H \; :  \only<5>{\phantom{\; \Worlds \fun {}}}\only<6>{\; \Worlds \fun} \mathrm{Pred}(\mathrm{MemSeg})
  \]
\end{overprint}
  \note{Memory changes over time, for instance new memory may be allocated.}
  \note{Allow worlds to evolve. New invariants can be added to handle freshly allocated memory.}
  \note{Safety of words monotone w.r.t. worlds (which makes it into a Kripke Logical relation).}
\end{frame}

@@

\begin{frame}[fragile]
  \frametitle{Local Capabilities}
  \texttt{f} is unknown code and \texttt{c} is a capability.
  \begin{lstlisting}
    f(c);
    f(1)
  \end{lstlisting}
  \begin{itemize}
  \item<2-> \texttt{c} global $\Rightarrow$ available in second invocation of \texttt{f}

    \note{when c global, second invocation in a world where c is safe}
  \item<3-> \texttt{c} local $\Rightarrow$ not available in second invocation of \texttt{f}
    \note{when c local, second invocation in a world where c is not (necessarily) safe}
    \note{c essentially revoked, so the invariants it relies on need not hold.}
  \end{itemize}

  \note{Two future world relations, one for all capabilities and one for non-local capabilities}
  \note{If c global, then second invocation must happen in public future world, so c valid.}
  \note{If c local, then second invocation may happen in a private future world.}
  \begin{onlyenv}<4->
  \begin{lemma}[Double monotonicity of value relation]
    \begin{itemize}
    \item If $\npair{w} \in \stdvr(W)$ and $W' \futurewk W$ then $\npair{w} \in
      \stdvr(W')$.
    \item If $\npair{w} \in \stdvr(W)$ and $W' \futurestr W$ and $w$ is
      not a local capability, then $\npair{w} \in \stdvr(W')$.
    \end{itemize}
  \end{lemma}
\end{onlyenv}
  \note{How does local/global capabilities affect all this.}
  \note{If we hand a global capability to untrusted code, then it may be stored in memory, so we will only be able to reinvoke that code if we can guarantee that those values are still valid. }
  \note{Formally, the worlds contain the invariants that the global capability depend on and reinvocation is only possible in future worlds where these invariants are respected.}
  \note{Local capabilities on the other provides a means to revoke capabilities. If we invoke untrusted code and give them a local capability, then they have no way to store it aside from the register file (and the stack), so when they return we can be sure that the local capability resides nowhere but the register file and the stack.}
  \note{If we clear the register file and stack for old local capabilities before we invoke the untrusted code again, then we have essentially revoked those local capabilities and we can safely invoke the untrusted code again with the assumption that it does not have access to the local capability from the first call.}
  \note{Formally, we allow this call to happen in a \emph{private future} world in which all global capabilities remain safe, but local capabilities don't.}
  \note{We also have a \emph{public future} world relation in which \emph{all} capabilities remain safe.}
  \note{Concretely, worlds may contain temporary regions on which local capabilities may depend. These regions persist in public future worlds which allows the local capabilities to rely on them. On the other hand, in private future worlds temporary regions may be revoked which is why local capabilities may not be safe in private future worlds.}

  \note{include comment about priv/pub future worlds + interplay with monotonicity?}
\end{frame}

\begin{frame}
  \frametitle{Fundamental Theorem of Logical Relations}
    \begin{itemize}
  \item General statement about the guarantees provided by the capability machine.
  \item Intuitively: any program is safe as long as it only has access to safe values.
  \end{itemize}
  \begin{theorem}[Fundamental theorem (simplified)]
      If 
      \[
        \npair{(\start,\addrend)} \in \readCond{}(\gl)(W)
      \]
      then
      \[
        \npair{((\exec,\gl),\start,\addrend,\addr)} \in \stder(W)
      \]
  \end{theorem}

  % Comments/present high-level perspective
  \note{readCond is the assumption that every thing in the interval [a,e] is safe to read.}
  \note{$\stder$ is safe to execute relation. That is, it will respect all the memory invariants.}
  \note{That is take an arbitrary capability. If it only has access to safe capabilities then it will preserve the invariants of the world.}
  \note{Remeber, dynamic checks = failing is concidered secure} 
\end{frame}

\begin{frame}[fragile]
  \frametitle{``Awkward Example''}
    \begin{tabular}{c}
    \begin{lstlisting}
let x = ref 0 in
  (*$\lambda$*)f.(x := 0;
      f();
      x := 1;
      f();
      assert(!x == 1))
    \end{lstlisting}
  \end{tabular}
\note{example known from the literature}
\note{even just in ML difficult as f can be the closure.}
\note{the assert can fail if the calls are not well-bracketed!}  
\note{the local state is difficult to handle as the closure and the context needs to be able to update the invariant for x in different ways. (closure can switch between 0 and 1 as it pleases, but context can transition only from 0 to 1.)}
\note{relies heavily on well-bracketedness}
\note{we have made a faithful translation and proved correctness (i.e., the assertion never fails).}
\note{more semantic statement of guarantees allows us to do this.}
\end{frame}

>>>>

\begin{frame}
  \frametitle{Worlds, Safe Values, and Step-Indexing}
\begin{columns}
  \begin{column}{0.5\textwidth}
    \begin{itemize}
      \note[item]<1-3>{Compared to standard assembly language, capabilities executing code has access to represent bound.}
    \item<1-> Capabilities represent bounds on executing code
      \note[item]<1-3>{That is, the capabilities the executing code has access to.}
      \note[item]<1-3>{What exactly is the bound on. In the system we have considered, no I/O, so memory invariants.}
      \note[item]<1-3>{Take what the program does into account - allows more fine-grained then simply ``read/write''}
    \item<2-> World, $W$
      \note[item]<2-3>{The collection of invariants is what we call the world (and it can bethought of as  a model of the memory) }
      \note[item]<2-3>{Here colored region represents invariant. A simple invariant could be a specific address contains 42.}
      \begin{itemize}
      \item Collection of invariants
      \end{itemize}
      \note[item]<3-7>{We are interested in all the words on the machine that preserve these invariants, so}
    \item<4-> Predicate for safe values w.r.t world, $\stdvr(W)$
      \note[item]<4-7>{We define a predicate with theses values. Whether a capability violates invariants obviously depend on the invariants, so the predicate has to be world-indexed.}
      \note[item]<4-7>{Let's see some examples:}
      \note[item]<5-7>{If the capability on the slide has \emph{read} authority, then it cannot violate the simple invariant.}
      \note[item]<6-7>{If the capability on the slide also has \emph{write} authority, then it can violate the invariant by simply overwriting that address with a different number. \emph{Not safe}.}
      \note[item]<7-7>{Generally speaking a capability that can read is safe when it only can read safe words. What happens when it is stored at an address that it has authority over itself?}
      \note[item]<7->{It is safe only if it can read only safe values which requires it to be safe.}
      \begin{itemize}[<9->]
      \item Recursively definition
        \note[item]<8->{Need to take a fixed-point. Made possible by step-indexing.}
        \note[item]<9->{Related to similar issue for languages with \emph{recursive types} and \emph{ML-like references}.}
      \end{itemize}
    \end{itemize}
\end{column}
\begin{column}{0.5\textwidth}
  \begin{tikzpicture}[scale=.5, every node={scale=.5}]
    \scope
      \clip (-.1,-.1) rectangle (6.1,15.1);
      \fill[fill=white] (0,0) rectangle (6,15);
      \draw (0,0) -- (0,15);
      \draw (6,0) -- (6,15);
    \endscope
    % \draw[->] (-2,0) -- node[midway,sloped,above] {stack grows upward} (-2,15);
    \draw (-1.5,0) node {};
    \draw (-1.5,15) node {};
    \foreach \x in {0.5,1,...,14}
    {
      \draw[fill=white] (0,\x) rectangle (6,\x+.5) node[pos=.5,color=teal] {};
    };

    % One capability
    \draw<1-7> [decorate,decoration={brace,amplitude=3pt,mirror,raise=-2pt},yshift=0pt]
    (6.25,2) -- (6.25,5) node[draw=black] (sp1) [black,midway,xshift=6pt] {}; % ,shape=coordinate,xshift=1pt
    \draw<1-4>  (6,7.25) edge[black,->,in=45] (sp1);

@@

    \fill<3-7>[black, fill opacity=0, text opacity=0.7] (0,4) rectangle (6,4.5) node[pos=.5] {\footnotesize 42};

    \draw<5> (6,7.25) edge[black,->,in=45] node[right] {\footnotesize \textsc{r}} (sp1)   ;
    \draw<6-7> (6,7.25) edge[black,->,in=45] node[right] {\footnotesize \textsc{rw}} (sp1)   ;
    \draw<7>[very thick,color=red] (6.5,6.5) -- (8.5,4.5)
    (6.5,4.5) -- (8.5,6.5);

    \draw<8-> [decorate,decoration={brace,amplitude=3pt,mirror,raise=-2pt},yshift=0pt]
    (6.25,2) -- (6.25,5) node[draw=black] (sp1) [black,midway,xshift=6pt] {}; % ,shape=coordinate,xshift=1pt
    \path<8->  (6,4.25) edge[black,out=30,in=135] node[above right]  {\footnotesize \textsc{r}} (8,5) 
              (8,5) edge[black,->,out=-45,in=45](sp1);

@@

  \begin{overprint}
\onslide<1-4>
  \begin{itemize}
\note[item]<1->{Like languages with ML-references, memory changes over time. Example, if we are in this memory with two invariants and more is memory allocation. World need to cope with this.}
  \item<1-> Memory evolves over time
    \note[item]<2->{Notion called \emph{future worlds}. Allow us to \emph{add invariants}.}
  \item<3-> Add invariants in future worlds
    \note[item]<3->{Memory may be repurposed. Satic invariants don't model this. Alle invariants have an internal state machine. \emph{State machine progress in future worlds}.}
  \item<4-> Invariants as state machines
    \note[item]<4->{Each state is associated with a predicate of all memories that respect the invariant.}
  \end{itemize}
  \onslide<5-6>
  \begin{itemize}
  \item<5-> Each state contains a predicate of accepted memory segments
    \note[item]<5->{ We want to express ``all memories with safe values''. World dependent, so the predicate needs to be world indexed. }
  \item<6-> World indexed
    \note[item]<6->{ Worlds with invariants with state machines with predicates that are world indexed - circular definition. }
    \note[item]<6->{ Resolved using standard techniques from the litterature (essentially advanced step-indexing). }
  \end{itemize}
  \[
    H \; :  \only<5>{\phantom{\; \Worlds \fun {}}}\only<6>{\; \Worlds \fun} \mathrm{Pred}(\mathrm{MemSeg})
  \]
\end{overprint}
  % \note{Memory changes over time, for instance new memory may be allocated.}
  % \note{Allow worlds to evolve. New invariants can be added to handle freshly allocated memory.}
  % \note{Safety of words monotone w.r.t. worlds (which makes it into a Kripke Logical relation).}
\end{frame}

@@

\begin{frame}[fragile]
  \frametitle{Local Capabilities}
  \note[item]<1>{Now consider how local capabilities affect all this.}
  \note[item]<1>{Consider this simple example, first f is called with capability c as an argument. Then f is called with unit.}
%  \note[item]<1>{What do we need to argue about c in the first call for it to be valid?}
  \note[item]<1>{What may we assume about c in the second invocation of f?}
  \note[item]<1>{Depends on c!}
  \note[item]<2-3>{(Cont) Depends on c!}
  \texttt{f} is unknown code and \texttt{c} is a capability.
  \begin{lstlisting}
    f(c);
    f(1)
  \end{lstlisting}
  \begin{itemize}
  \item<2-> \texttt{c} global $\Rightarrow$ available in second invocation of \texttt{f}
    \note[item]<2-3>{If \emph{c is global}, then it can be \emph{stored on the heap}, so it needs to remain safe for the remainder of the execution. When f is invoked \emph{c must still be safe}.}
%    \note{when c global, second invocation in a world where c is safe}
  \item<3-> \texttt{c} local $\Rightarrow$ not available in second invocation of \texttt{f}
    \note[item]<3>{If \emph{c local}, then \emph{CC dictates clear} all the places c may reside, so in the second invocation \emph{c need not remain safe}.}
%    \note{when c local, second invocation in a world where c is not (necessarily) safe}
%    \note{c essentially revoked, so the invariants it relies on need not hold.}
  \end{itemize}
  \note[item]<3>{Need two future world relations. In both, global capabilities must remain safe. In one local capabilities need not. We have public and private future world relation.}
  \note[item]<4->{This double monotonicity lemma expresses the assumptions we can make, namely in \emph{public future world all capabilities} remain valid and in \emph{private future worlds local capabilities need not remain valid}.}
  \note[item]<4->{In the example this means that if \emph{c is local}, then it is okay to invoke f a second time in a \emph{private future world} as c need not be safe anymore. }
  \note[item]<4>{On the other hand, if \emph{c is global}, then the invokation of f must be in a \emph{public future world}, so c remains safe.}
%  \note{Two future world relations, one for all capabilities and one for non-local capabilities}
%  \note{If c global, then second invocation must happen in public future world, so c valid.}
%  \note{If c local, then second invocation may happen in a private future world.}
\onslide<4->{
  \begin{lemma}[Double monotonicity of safety predicate]
    \begin{itemize}
    \item If $\npair{w} \in \stdvr(W)$ and $W' \futurewk W$ then $\npair{w} \in
      \stdvr(W')$.
    \item If $\npair{w} \in \stdvr(W)$ and $W' \futurestr W$ and $w$ is
      not a local capability, then $\npair{w} \in \stdvr(W')$.
    \end{itemize}
  \end{lemma}}
\note[item]<4>{Related to public private future worlds, state machines with pub/priv transitions}
%  \note{How does local/global capabilities affect all this.}
%  \note{If we hand a global capability to untrusted code, then it may be stored in memory, so we will only be able to reinvoke that code if we can guarantee that those values are still valid. }
%  \note{Formally, the worlds contain the invariants that the global capability depend on and reinvocation is only possible in future worlds where these invariants are respected.}
%  \note{Local capabilities on the other provides a means to revoke capabilities. If we invoke untrusted code and give them a local capability, then they have no way to store it aside from the register file (and the stack), so when they return we can be sure that the local capability resides nowhere but the register file and the stack.}
%  \note{If we clear the register file and stack for old local capabilities before we invoke the untrusted code again, then we have essentially revoked those local capabilities and we can safely invoke the untrusted code again with the assumption that it does not have access to the local capability from the first call.}
%  \note{Formally, we allow this call to happen in a \emph{private future} world in which all global capabilities remain safe, but local capabilities don't.}
%  \note{We also have a \emph{public future} world relation in which \emph{all} capabilities remain safe.}
%  \note{Concretely, worlds may contain temporary regions on which local capabilities may depend. These regions persist in public future worlds which allows the local capabilities to rely on them. On the other hand, in private future worlds temporary regions may be revoked which is why local capabilities may not be safe in private future worlds.}

%  \note{include comment about priv/pub future worlds + interplay with monotonicity?}
\end{frame}

\begin{frame}
  \frametitle{Fundamental Theorem of Logical Relations}
    \begin{itemize}
  \item General statement about the guarantees provided by the capability machine.
  \item Intuitively: any program is safe as long as it only has access to safe values.
  \end{itemize}
  \begin{theorem}[Fundamental theorem (simplified)]
      If 
      \[
        \npair{(\start,\addrend)} \in \readCond{}(\gl)(W)
      \]
      then
      \[
        \npair{((\exec,\gl),\start,\addrend,\addr)} \in \stder(W)
      \]
  \end{theorem}
  \note[item]<1->{Now for the formal statement about guarantees}
  \note[item]<1->{readCond: only safe values in the interval.}
  \note[item]<1->{E-relation: when capability used as the pc with register file and memory respecting the world, then the execution respects the memory invariants.}

  \note[item]<1->{In other words, take an arbitrary capability that only has access to safe values, then the memory invariants are preserved when we use it for execution.}
  \note[item]<1->{The instructions it execute don't matter only the authority it can use.}
  % Comments/present high-level perspective
  % \note{readCond is the assumption that every thing in the interval [a,e] is safe to read.}
  % \note{$\stder$ is safe to execute relation. That is, it will respect all the memory invariants.}
  % \note{That is take an arbitrary capability. If it only has access to safe capabilities then it will preserve the invariants of the world.}
  % \note{Remeber, dynamic checks = failing is concidered secure} 
\end{frame}

\begin{frame}[fragile]
  \frametitle{``Awkward Example''}
    \begin{tabular}{c}
    \begin{lstlisting}
let x = ref 0 in
  (*$\lambda$*)f.(x := 0;
      f();
      x := 1;
      f();
      assert(!x == 1))
    \end{lstlisting}
  \end{tabular}
  \note[item]{known from the litterature}
  \note[item]{in ML difficult to reason about as callback f can be the closure it self. (so x can be either )}
  \note[item]{assert may fail if calls not well-bracketed}
  \note[item]{can do more things to attack well-bracketedness low-level. Context need not follow CC, so well-behaved code cannot rely on behavior of untrusted code. }
  \note[item]{ We have proven a faithful translation correct. That is the assert never fails. Notice, dynamic checks, so machine can fail, but we set it up, so we can distinguish this from assertion failure.}
  \note[item]{ More semantic statement of guarantees allow us to prove it. }
% \note{example known from the literature}
% \note{even just in ML difficult as f can be the closure.}
% \note{the assert can fail if the calls are not well-bracketed!}  
% \note{the local state is difficult to handle as the closure and the context needs to be able to update the invariant for x in different ways. (closure can switch between 0 and 1 as it pleases, but context can transition only from 0 to 1.)}
% \note{relies heavily on well-bracketedness}
% \note{we have made a faithful translation and proved correctness (i.e., the assertion never fails).}
% \note{more semantic statement of guarantees allows us to do this.}
\end{frame}
